---
title: "Customer Segmentation"
author: "Trung Le"
date: "February 10, 2017"
output: html_document
---

## Overview

It is very important nowadays for company to understand their customer, know the value of each customer in their business as customers are not all equal. Knowing which customers are the most valueable, company can build marketing strategy to increase thier satisfaction with lower cost and more effectively. Moreover, understanding customers behavior, company can increase revenue by sending the marketing massages to the right customer at the right time in the right place. This assginment will cover 2 topics: customer segmentation and customer behavior.

## Load the data

Install all the packages requirement and read the data

```{r}
library(dplyr)
library(gplots)
library(ggplot2)
library(cluster)
library(devtools)
install_github("hoxo-m/easyRFM")
library(easyRFM)
install_github("Rdatatable/data.table")
library(data.table)
library(cluster)
library(fpc)
dat <- read.csv("Superstore.csv")
```
## Tidy the data

There are many variables that are not neccessary for our purposes, so remove them. Then convert variables and solve NA observation

#### Remove variable and convert variables
```{r}
cust <- dat[,c(1, 6, 2:3, 14:22)]
cust$Customer.ID <- as.character(cust$Customer.ID)
cust$Order.ID <- as.character(cust$Order.ID)
cust$Order.Date <-as.Date(cust$Order.Date, format = "%m/%d/%y")
cust$Product.ID <- as.character(cust$Product.ID)
cust$Product.Name <- as.character(cust$Product.Name)
cust$Sales <- as.numeric(as.character(gsub(",","",cust$Sales)))
cust$Profit <- as.numeric(as.character(gsub(",","",cust$Profit)))
```

Structure of the data
```{r}
str(dat)
```

#### Treating missing observations

There are 668 missing observation which is 1.3% of the total data. So we delete these missing observations
```{r}
sum(is.na(cust$Sales))
sum(is.na(cust$Profit))
cust <- na.omit(cust)
names(cust$Sub.Category)
```
## Feature new variables
In order to understand customer behavior based on products they bought in the past, it is important to create new variable of product categories. We use packages 'dplyr' to feature these variable and then combines all of them into sub.cat data

```{r}
# Phones
Phones <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Phones = as.numeric(ifelse(Sub.Category == "Phones", 1, 0)), Phones_Num = Phones*Quantity)

Phones <- Phones %>% group_by(Customer.ID) %>% select(Customer.ID, Phones_Num) %>% summarise(Phones_Num = sum(Phones_Num))

# Chairs
Chairs <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Chairs = as.numeric(ifelse(Sub.Category == "Chairs", 1, 0)), Chairs_Num = Chairs*Quantity)
                                                                                                    
Chairs <- Chairs %>% group_by(Customer.ID) %>% select(Customer.ID, Chairs_Num) %>% summarise(Chairs_Num = sum(Chairs_Num))                                                                                                 
# Tables
Tables <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Tables = as.numeric(ifelse(Sub.Category == "Tables", 1, 0)), Tables_Num = Tables*Quantity)

Tables <- Tables %>% group_by(Customer.ID) %>% select(Customer.ID, Tables_Num) %>% summarise(Tables_Num = sum(Tables_Num))       

# Copiers
Copiers <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Copiers = as.numeric(ifelse(Sub.Category == "Copiers", 1, 0)), Copiers_Num = Copiers*Quantity)

Copiers <- Copiers %>% group_by(Customer.ID) %>% select(Customer.ID, Copiers_Num) %>% summarise(Copiers_Num = sum(Copiers_Num))       

# Bookcases
Bookcases <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Bookcases = as.numeric(ifelse(Sub.Category == "Bookcases", 1, 0)), Bookcases_Num = Bookcases*Quantity)

Bookcases <- Bookcases %>% group_by(Customer.ID) %>% select(Customer.ID, Bookcases_Num) %>% summarise(Bookcases_Num = sum(Bookcases_Num))       

# Art
Art <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Art = as.numeric(ifelse(Sub.Category == "Art", 1, 0)), Art_Num = Art*Quantity)

Art <- Art %>% group_by(Customer.ID) %>% select(Customer.ID, Art_Num) %>% summarise(Art_Num = sum(Art_Num))       

# Appliances
Appliances <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Appliances = as.numeric(ifelse(Sub.Category == "Appliances", 1, 0)), Appliances_Num = Appliances*Quantity)

Appliances <- Appliances %>% group_by(Customer.ID) %>% select(Customer.ID, Appliances_Num) %>% summarise(Appliances_Num = sum(Appliances_Num))       

# Storage
Storage <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Storage = as.numeric(ifelse(Sub.Category == "Storage", 1, 0)), Storage_Num = Storage*Quantity)

Storage <- Storage %>% group_by(Customer.ID) %>% select(Customer.ID, Storage_Num) %>% summarise(Storage_Num = sum(Storage_Num))       

# Fasteners
Fasteners <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Fasteners = as.numeric(ifelse(Sub.Category == "Fasteners", 1, 0)), Fasteners_Num = Fasteners*Quantity)

Fasteners <- Fasteners %>% group_by(Customer.ID) %>% select(Customer.ID, Fasteners_Num) %>% summarise(Fasteners_Num = sum(Fasteners_Num))       

# Machines
Machines <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Machines = as.numeric(ifelse(Sub.Category == "Machines", 1, 0)), Machines_Num = Machines*Quantity)

Machines <- Machines %>% group_by(Customer.ID) %>% select(Customer.ID, Machines_Num) %>% summarise(Machines_Num = sum(Machines_Num))       

# Accessories
Accessories <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Accessories = as.numeric(ifelse(Sub.Category == "Accessories", 1, 0)), Accessories_Num = Accessories*Quantity)

Accessories <- Accessories %>% group_by(Customer.ID) %>% select(Customer.ID, Accessories_Num) %>% summarise(Accessories_Num = sum(Accessories_Num))       

# Furnishings
Furnishings <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Furnishings = as.numeric(ifelse(Sub.Category == "Furnishings", 1, 0)), Furnishings_Num = Furnishings*Quantity)

Furnishings <- Furnishings %>% group_by(Customer.ID) %>% select(Customer.ID, Furnishings_Num) %>% summarise(Furnishings_Num = sum(Furnishings_Num))       

# Binders
Binders <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Binders = as.numeric(ifelse(Sub.Category == "Binders", 1, 0)), Binders_Num = Binders*Quantity)

Binders <- Binders %>% group_by(Customer.ID) %>% select(Customer.ID, Binders_Num) %>% summarise(Binders_Num = sum(Binders_Num))       

# Labels 
Labels <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Labels = as.numeric(ifelse(Sub.Category == "Labels", 1, 0)), Labels_Num = Labels*Quantity)

Labels <- Labels %>% group_by(Customer.ID) %>% select(Customer.ID, Labels_Num) %>% summarise(Labels_Num = sum(Labels_Num))       

# Supplies
Supplies <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Supplies = as.numeric(ifelse(Sub.Category == "Supplies", 1, 0)), Supplies_Num = Supplies*Quantity)

Supplies <- Supplies %>% group_by(Customer.ID) %>% select(Customer.ID, Supplies_Num) %>% summarise(Supplies_Num = sum(Supplies_Num))       

# Envelopes
Envelopes <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Envelopes = as.numeric(ifelse(Sub.Category == "Envelopes", 1, 0)), Envelopes_Num = Envelopes*Quantity)

Envelopes <- Envelopes %>% group_by(Customer.ID) %>% select(Customer.ID, Envelopes_Num) %>% summarise(Envelopes_Num = sum(Envelopes_Num))   

# Paper
Paper <- cust %>% group_by(Customer.ID) %>% select(Customer.ID, Sub.Category, Quantity) %>% mutate(Paper = as.numeric(ifelse(Sub.Category == "Paper", 1, 0)), Paper_Num = Paper*Quantity)

Paper <- Paper %>% group_by(Customer.ID) %>% select(Customer.ID, Paper_Num) %>% summarise(Paper_Num = sum(Paper_Num))       

# combine all sub.category
sub.cat <- cbind(Storage, Phones, Machines, Labels, Furnishings, Fasteners, Envelopes, Copiers, Chairs, Bookcases, Binders, Art, Appliances, Accessories, Tables, Supplies, Paper)

sub.cat <- sub.cat[,c(1,2,4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34)]
rm(Accessories, Appliances, Art, Binders, Bookcases, Chairs, Copiers, Envelopes, Fasteners, Furnishings, Labels, Machines, Paper, Phones, Storage, Supplies, Tables)
```

## RFM model
We use RFM model to get groups of customer: leaved customers, leaving customers, good customers, new customers

```{r}
RFM <- rfm_auto(cust, id="Customer.ID", payment="Sales", date="Order.Date")
head(RFM$rfm)

# based on 3 variables, R, F, M, we create 4 groups of customers
leaved_customers <- RFM$get_sliced_rfm(R_slice=1:2)
leaving_customers <- RFM$get_sliced_rfm(R_slice=3)
good_customers <- RFM$get_sliced_rfm(R_slice=4:5, F_slice=4:5)
new_customers <- RFM$get_sliced_rfm(R_slice=4:5, F_slice=1:2)

rfm.score <- RFM$rfm
str(rfm.score)
rfm.score$score <- rfm.score$RecencyClass+rfm.score$FrequencyClass+rfm.score$MonetaryClass
rfm.score$RecencyClass <- as.factor(rfm.score$RecencyClass)
rfm.score$FrequencyClass <- as.factor(rfm.score$FrequencyClass)

# create baloon plot
balloonplot(rfm.score$RecencyClass, rfm.score$FrequencyClass, rfm.score$MonetaryClass)
```
## Clustering technique
We will do clustering based on product category for each group of customer to understand the purchasing behavior of each group. Based on that we can set marketing strategy for each group of customers.
Now we need to merge each group of customers with sub.cat data to do clustering. In order to do that, we create a variable named Cs.Type which:
0 = leaved customers
1 = leaving customers
2 = Good customers
3 = new customers
4 = monitoring

```{r}
sub.cat$Cs.Type <- 4
sub.cat$Cs.Type[sub.cat$Customer.ID %in% good_customers$Customer.ID] = 2
sub.cat$Cs.Type[sub.cat$Customer.ID %in% leaved_customers$Customer.ID] = 0
sub.cat$Cs.Type[sub.cat$Customer.ID %in% leaving_customers$Customer.ID] = 1
sub.cat$Cs.Type[sub.cat$Customer.ID %in% new_customers$Customer.ID] = 3

# filtering to get product categories data of each group
good.clust <- sub.cat[sub.cat$Cs.Type == 2,]
leaved.clust <- sub.cat[sub.cat$Cs.Type == 0,]
leaving.clust <- sub.cat[sub.cat$Cs.Type == 1,]
new.clust <- sub.cat[sub.cat$Cs.Type == 3,]
```
#### K-mean clustering

##### Good customer group
```{r}
set.seed(1)
good.clust <- good.clust[,-c(1,19)]
# finding number of clusters
wss <- (nrow(good.clust)-1)*sum(apply(good.clust,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(good.clust, centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")

# pick k = 4
k1 <- kmeans(good.clust, 4)
k1$size
k1$centers

# plot good customers cluster
plotcluster(good.clust, k1$cluster)
```
##### Leaved customer group
```{r}
leaved.clust <- leaved.clust[,-c(1,19)]
# finding number of clusters
wss <- (nrow(leaved.clust)-1)*sum(apply(leaved.clust,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(leaved.clust, centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")

# pick k = 7
k2 <- kmeans(leaved.clust, 7)
k2$size
k2$centers

# plot good customers cluster
plotcluster(leaved.clust, k2$cluster)

```
##### Leaving customer group
```{r}
leaving.clust <- leaving.clust[,-c(1,19)]
# finding number of clusters
wss <- (nrow(leaving.clust)-1)*sum(apply(leaving.clust,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(leaving.clust, centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")

# pick k = 6
k3 <- kmeans(leaving.clust, 6)
k3$size
k3$centers

# plot good customers cluster
plotcluster(leaving.clust, k3$cluster)
```
##### New customer group
```{r}
new.clust <- new.clust[,-c(1,19)]
# finding number of clusters
wss <- (nrow(new.clust)-1)*sum(apply(new.clust,2,var))
for (i in 2:15) wss[i] <- sum(kmeans(new.clust, centers=i)$withinss)
plot(1:15, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")

# pick k = 10
k4 <- kmeans(new.clust, 10)
k4$size
k4$centers

# plot good customers cluster
plotcluster(new.clust, k4$cluster)
```

